### 2008-12-20 talk/ecoop2009/clibackend.tex
remove the jitstrategy section and integrate its contents into the introduction.
eh. not that many references were left to the rainbow interp anyway. die die
(arigato around) add an XXX
expand on the XXX
several small fixes.  Some rewording in the cli backend section

\section{CLI backend for Rainbow interpreter}

@@

@@

As already explained, \dacom{I guess flexswitches will be introduced
  in the previous sect.} \emph{flexswitch} is one of the key
concepts allowing the JIT compiler generator to produce code which can
be incrementally specialized and compiled at run time.

A flexswitch is a special kind of switch which can be dynamically
extended with new cases; intuitively, its behavior can be described
well in terms of flow graphs. Indeed, a flexswitch can be considered 
as a special flow graph block where links to newly created blocks are
dynamically added whenever new cases are needed.

@@

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.
In this way, whereas flow graphs without flexswitches are translated
to a single method, the translation of flow graphs which can dynamically grow because of
flexswitches will be scattered over several methods.
Summarizing, the backend behaves in the following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. Each collection contains at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph. Among these blocks
  there always exists an initial block whose input arguments 
  might be passed as arguments of the method; however, for
  implementation  reasons (see the details below) the input variables
  of all blocks (including the initial one)
  are implemented as local variables of the method. 
\end{itemize}

When  a new case is added to a flexswitch, new blocks are generated
and translated by the backend in a new single method pointed
by a delegate \footnote{\emph{Delegates} are the .NET equivalent of function pointers}
 of  which is stored in the code implementing the flexswitch,
so that the method can be invoked later.

@@

A link is called \emph{internal} if it connects two blocks implemented
by the same method,
 \emph{external} otherwise.

Following an internal link would  not be difficult in IL bytecode: a jump to
the corresponding code fragment in the same method can be emitted 
to execute the new block, whereas the appropriate local variables can be
used for passing arguments. 
Also following an external link whose target is an initial block could
be easily implemented, by just invoking the corresponding method.

What cannot be easily implemented in CLI is following an external link
whose target is not an initial block; consider, for instance, the
outgoing link of the block dynamically added in the right-hand side
picture of Figure~\ref{flexswitch-fig}. How it is possible to pass the
right arguments to the target block?

To solve this problem every method contains a special code, called
\emph{dispatcher}; whenever a method is invoked, its dispatcher is
executed first\footnote{The dispatcher should not be
confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high 2 bytes \dacom{word was meant as a fixed-sized group of bits} of a block id constitute the id of the method to which the block
belongs, whereas the low 2 bytes constitute a progressive number univocally identifying
each block implemented by the method.

@@

Therefore, the only solution we came up with is defining a class
\lstinline{InputArgs} for passing sequences of arguments whose length
and type is variable.
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
public class InputArgs {
  public int[] ints;
  public float[] floats;
  public object[] objs;
  ...
}
\end{lstlisting}
\end{small}
Unfortunately, with this solution passing arguments to external links
becomes quite slow:
\begin{itemize}
\item When writing arguments, array re-allocation may be needed in
  case the number of arguments exceeds the dimension of the
  array. Furthermore the VM will always perform bound-checks, even
  when the size is explicitly checked in advance;

>>>>

\section{The CLI backend for the Generated JIT}
\label{sec:clibackend}

@@

\anto{XXX: add an overview section of the backend, else we only talk about
  flexswitches}
\cfbolz{it should maybe also have a nice introductory paragraph how the
interface between the JIT works?}

@@

As already explained, \emph{flexswitch} is one of the key
concepts allowing the JIT compiler generator to produce code which can
be incrementally specialized and compiled at run time.

A flexswitch is a special kind of switch which can be dynamically
extended with new cases.  Intuitively, its behavior can be described
well in terms of flow graphs: a flexswitch can be considered 
as a special flow graph block where links to newly created blocks are
dynamically added whenever new cases are needed.

@@

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.

It is important to underline the difference between flow graphs and a methods:
the first are the logical unit of code as seen by the JIT compiler, each of
them being concretely implemented by \emph{one or more} methods.

In this way, whereas flow graphs without flexswitches are translated to a
single method, the translation of \emph{growable} flow graphs will be
scattered over several methods.  Summarizing, the backend behaves in the
following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. Each collection contains at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph.
\end{itemize}

When a new case is added to a flexswitch, the backend generates the new blocks
into a new single method.  The newly created method is pointed by a
delegate \footnote{\emph{Delegates} are the .NET equivalent of function
  pointers} stored in the flexswitch, so that it can be invoked later when
needed.

@@

A link is called \emph{internal} if it connects two blocks contained
in the same method,
 \emph{external} otherwise.

Following an internal link is easy in IL bytecode: a jump to
the corresponding code fragment in the same method can be emitted 
to execute the new block, whereas the appropriate local variables can be
used for passing arguments.

Following an external link whose target is an initial block could also
be easily implemented, by just invoking the corresponding method.
What cannot be easily implemented in CLI is following an external link
whose target is not an initial block; consider, for instance, the
outgoing link of the block dynamically added in the right-hand side
picture of Figure~\ref{flexswitch-fig}. How it is possible to jump in 
the middle of a method?

To solve this problem every method contains a special code, called
\emph{dispatcher}: whenever a method is invoked, its dispatcher is
executed first\footnote{The dispatcher should not be
confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high 2 bytes of a block id constitute the \emph{method id}, which 
univocally identifies a method in a graph, whereas the low 2 bytes constitute
a progressive number univocally identifying a block inside each method.

@@

Therefore, the solution we came up with is defining a class
\lstinline{InputArgs} for passing sequences of arguments whose length
and type is variable.
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
public class InputArgs {
  public int[] ints;
  public float[] floats;
  public object[] objs;
  ...
}
\end{lstlisting}
\end{small}
Unfortunately, with this solution passing arguments to external links
becomes quite slow:
\begin{itemize}
\item When writing arguments, array re-allocation may be needed in
  case the number of arguments exceeds the dimension of the
  array. Furthermore the VM will always perform bound-checks, even
  when the size is explicitly checked in advance;

### 2008-12-20 talk/ecoop2009/tlc.tex
make a figure
several small fixes.  Some rewording in the cli backend section

\begin{itemize}
\item \textbf{Stack manipulation}: standard operations to manipulate the
  stack, such as \lstinline{POP}, \lstinline{PUSHARG}, \lstinline{SWAP}, etc.
\item \textbf{Flow control} to do conditional (\lstinline{BR_COND}) and
  unconditional (\lstinline{BR}) jumps.
\item \textbf{Arithmetic}: numerical operations on integers, like
  \lstinline{ADD}, \lstinline{SUB}, etc.
\item \textbf{Comparisons} like \lstinline{EQ}, \lstinline{LT},
  \lstinline{GT}, etc.
\item \textbf{Object-oriented}: operations on objects: \lstinline{NEW},
  \lstinline{GETATTR}, \lstinline{SETATTR}, \lstinline{SEND}.
\item \textbf{List operations}: \lstinline{CONS}, \lstinline{CAR},
  \lstinline{CDR}.
\end{itemize}

@@

As we said above, TLC exists only at bytecode level; to ease the development
of TLC programs, we wrote an assembler that generates TLC bytecode. The
following example shows a simple program that computes the absolute value of
the given integer:

\begin{lstlisting}
main:             # stack: []
    PUSHARG       #        [n]
    PUSH 0        #        [n, 0]
    LT            #        [0 or 1]
    BR_COND neg

@@

neg:
    PUSH 0        #        [0]
    PUSHARG       #        [0,n]
    SUB           #        [-n]
    RETURN
\end{lstlisting}

>>>>

\begin{itemize}
\item \textbf{Stack manipulation}: standard operations to manipulate the
  stack, such as \lstinline{POP}, \lstinline{PUSHARG}, \lstinline{SWAP}, etc.
\item \textbf{Flow control} to do conditional (\lstinline{BR_COND}) and
  unconditional (\lstinline{BR}) jumps.
\item \textbf{Arithmetic}: numerical operations on integers, like
  \lstinline{ADD}, \lstinline{SUB}, etc.
\item \textbf{Comparisons} like \lstinline{EQ}, \lstinline{LT},
  \lstinline{GT}, etc.
\item \textbf{Object-oriented} operations: \lstinline{NEW},
  \lstinline{GETATTR}, \lstinline{SETATTR}, \lstinline{SEND}.
\item \textbf{List operations}: \lstinline{CONS}, \lstinline{CAR},
  \lstinline{CDR}.
\end{itemize}

@@

As we said above, TLC exists only at bytecode level; to ease the development
of TLC programs, we wrote an assembler that generates TLC bytecode. Figure \ref{fig:tlc-abs}
shows a simple program that computes the absolute value of
the given integer.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}
main:             # stack: []
    PUSHARG       #        [n]
    PUSH 0        #        [n, 0]
    LT            #        [0 or 1]
    BR_COND neg

@@

neg:
    PUSH 0        #        [0]
    PUSHARG       #        [0,n]
    SUB           #        [-n]
    RETURN
\end{lstlisting}
\caption{The TLC bytecode for computing the absolute value of a function}
\label{fig:tlc-abs}
\end{center}
\end{figure}
